---
title: "Analysis on LendingClub's Loan Data"
author: "Uyemaa Gantulga, Ayush Meshram, Aakash Singh, and Melissa Yago"
date: "10/20/24"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=F}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

# Overview

Since 2006, LendingClub has been a U.S. financial services company that has facilitated loan contracts. Our research will focus on LendingClub's loan data that consist of approved and rejected loans between 2013-2018. To understand LendingClub’s clientele, we will analyze borrower and loan characteristics to assess risk levels, determine loan acceptance, and understand what factors influence whether accepted loans will be fully repaid. To accomplish this, we will ask the following SMART questions:


1. How do the loan features of term, amount, and interest rate individually influence the likelihood of a loan being charged off (lender has accepted the loan as unlikely to recover) versus being fully paid? 

2. How do variables such as dti, annual_inc, int_rate, and fico_range impact the likelihood of a loan being "fully paid" or "charged off" year over year from 2014 to 2016? What trends can be identified in the data, and how do these variables differ between the two loan statuses over time?

3. Do the loan grades provided to each customer correlate to their loan repayment behavior based on income and loan status? 

4. Is there a relationship between employment length and the amount requested for rejected loans and how does it compare to the customers that had accepted loans?

# Data Dictionary

1. Funded Amount: The total amount of the loan to be paid off.

2. Loan Status: Whether the loan has been fully paid or has been charged off. A loan that is fully paid is defined as a loan where the total payment has been met. A charge off status is defined as the lender accepting the loan as uncollectible. This differs from the default status in which payments are late but the lender expects to collect what the borrower still owes.  

3. Term: How long the borrower has to pay off the loan (Either 36 or 60 Months).

4. Interest Rate: The percentage of the loan amount that the lender charges the borrower for borrowing money. 

5. DTI (Debt to Income Ratio): A measure used to evaluate a borrower's financial situation by comparing their debt payments to their gross monthly income. 

6. Annual Income: The dollar amount the customer earns on a yearly basis. 

7. FICO Range: A measure of creditworthiness used by lenders to assess a borrower's ability to repay a loan. 

8. Loan Grades: Classifies the risk level of a loan from A (less risky) to G (risky). 

9. Employment Length: The amount in years the customer has been employed. 


# Question 1: How do the loan features of term, amount, and interest rate individually influence the likelihood of a loan being charged off versus being fully paid? 

This question focuses on assessing risk levels and looking into trends of when the company can expect to see a loan charged off (lender has accepted the loan as unlikely to recover) or being paid off. 

Assumptions:
In terms of loan status, we are only analyzing the "Fully Paid" and "Charge Off" status.

## Data Types & Data Cleaning

For this question, we will be looking at funded amount, loan status, term, and interest rate. 

Based on this data, the funded amount and interest rate are data types that are numerical. On the other hand, loan status and term have been converted to factors and are categorical data types. While loan status logically makes sense to be categorical, term has been converted to a factor since there were only two types of loan terms (36 or 60 months) which can be seen as short term vs. long term when comparing the likelihood of default across loans. Term in this case will be viewed as nonlinear with interest rates and funded amount affecting risk with linearity.

To retrieve this information, data cleaning was done on the accepted loans dataset that initially contained 2,157,894 million observations. After filtering, the total observations stand at 1,245,285 observations.    



```{r, echo=TRUE}
#Load Library
library(dplyr)

#Read the CSV
df <- read.csv("cleaned_accepted_2013_to_2018Q4.csv")

#Subset for Specific Columns
f_df<- df[, c("funded_amnt", "loan_status","term","int_rate")]

#Filter for specific loan status
df_filter <- f_df %>%
  filter(loan_status %in% c("Charged Off", "Fully Paid"))

#Convert loan status and term as factors
df_filter$loan_status <- as.factor(df_filter$loan_status)
df_filter$term <- as.factor(df_filter$term)


#Select Relevant Columns
df_final <-df_filter %>%
  select(funded_amnt, term, int_rate, loan_status)

#Check the Dataframe output
str(df_final)



```

## Normality of Loan Amount & Interest Rate

### QQ Plot of Loan Amount

```{r, echo=TRUE}
library(ezids)
amount_plot = outlierKD2(df_final, funded_amnt, rm = TRUE, qqplt = TRUE, histogram = FALSE)

```

The loan amount data does not appear to follow a normal distribution based on the QQ plot, as there are significant deviations from the reference line, particularly at the tails. The lower quantiles fall below the line, while the upper quantiles rise far above it, indicating the presence of heavy tails. Additionally, the stepped line in the middle suggests that the data isn't evenly distributed. Even with the outliers removed, there are still extreme deviations at both ends therefore showing that this is not a normal distribution.    

### QQ Plot of Interest Rate


```{r, echo=TRUE}
library(ezids)
rate_plot = outlierKD2(df_final, int_rate, rm = TRUE, qqplt = TRUE, histogram = FALSE)

```

The interest rate also shows similar trends. There are major deviations from the reference line, especially in the upper tail, indicating that the interest rate distribution is not normal when outliers are present. Even with the outliers removed, the QQ plot appears much more linear showing significant improvement towards normality, though there are deviations that remain in the tails. 

Overall, both loan amount and interest rate distributions are not normal when outliers are included. While showing improvement in normality when removing outliers, deviations still occur at the extremes. As a result, these datasets don't follow normality due to these deviations. 

## Descriptive Statistics

Initial statistics summarizes data that can help in identifying patterns, trends, and outliers. To demonstrate this, loan status will be analyzed against each loan feature (term, amount and interest rate) to see any relationships and patterns.  

### Relationship between Loan Status and Loan Term

```{r, echo=TRUE}
# Initialize Library
library(ggplot2)

# Contingency Table (Correlation Analysis)
c_table <- table(df_final$term, df_final$loan_status)
cat("Contingency Table: Loan Status by Term\n\n")
print(c_table)

#Contingency Table by Proportions
prop_table <-prop.table(c_table) *100
cat("Contingency Table: Loan Status by Term Proportions (by %)\n\n")
print(prop_table)

# Stacked Bar Chart
ggplot(df_final, aes(x = term, fill = loan_status)) + 
  geom_bar(position = "stack") +
  labs(x="Loan Term", y="Number of Loans",title="Stacked Bar Chart of Loan Status by Term" ) +
  scale_fill_manual(values = c("navy", "lightblue"))

  
```

Via the first contingency table, 36 month loans have a higher absolute number of both fully paid and charged off loans compared to 60 month loans. Proportionally, 36 month loans have a higher rate of being charged off (12.32%) than 60 month loans (8.01%).

When visually shown on a stacked bar chart, you'll notice that shorter terms are more likelier to get charged off than loans with a longer term.  

Overall, shorter loan terms are likelier to result in a charge off and appear to influence loan status. 


### Relationship between Loan Status and Loan Amount

```{r, echo=TRUE}
# Descriptive Stats of Loan Amount Separated by Loan Status

amnt_stats <- df_final %>%
  group_by(loan_status) %>%
  summarise(
    Mean = mean(funded_amnt, na.rm = TRUE),
    Median = median(funded_amnt, na.rm = TRUE),
    SD = sd(funded_amnt, na.rm = TRUE),
    Q25 = quantile(funded_amnt, 0.25, na.rm = TRUE),
    Q50 = quantile(funded_amnt, 0.50, na.rm = TRUE),
    Q75 = quantile(funded_amnt, 0.75, na.rm = TRUE),
    Min = min(funded_amnt, na.rm = TRUE),
    Max = max(funded_amnt, na.rm = TRUE)
  )

cat("Descriptive Stats of Loan Amounts Separated by Loan Status)\n\n")
print(amnt_stats)


```

```{r, echo=TRUE}
# Box Plot Comparing Loan Amounts by Loan Status
ggplot(df_final, aes(x = loan_status, y=funded_amnt, fill = loan_status)) + 
  geom_boxplot() + 
  geom_boxplot( colour="black", outlier.colour="blue", outlier.shape=8, outlier.size=4) +
  labs(title="Boxplot of Loan Amounts by Loan Status", x="Loan Status", y = "Loan Amount ($)")+ 
  scale_fill_manual(values = c("navy", "lightblue"))

```

Based on the statistics, the charged off loans having a higher mean and median loan amount, this can suggest that larger loan amounts may be associated with a higher risk of being charged off. Charged off loans also have a slightly higher standard deviation showing for more variability in the amounts.Lastly, both loan statuses show the same minimum and maximum but that does not necessarily mean that the distributions are identical.

Based on the boxplot, one commonality is the several outliers that exist around $40,000 for both loan statuses. However, this is not indicative that it is caused by status and requires further investigation through these high amounts being affected by other factors. Given the IQR of both statuses, there is not significant variability. You can however tell based on the IQR that the charged off loans have a slightly higher loan amounts compared to the fully paid loans that have a slightly lower loan amount.

Overall, larger loan amounts may be associated with a higher charged off risk. However holistically, loan amount alone does not appear to have a strong influence on whether a loan will be charged off or fully paid, given overlapping distributions. 


### Relationship Between Loan Status and Interest Rate

```{r, echo=TRUE}
# Descriptive Stats of Interest Rate Separated by Loan Status

rate_stats <- df_final %>%
  group_by(loan_status) %>%
  summarise(
    Mean = mean(int_rate, na.rm = TRUE),
    Median = median(int_rate, na.rm = TRUE),
    SD = sd(int_rate, na.rm = TRUE),
    Q25 = quantile(int_rate, 0.25, na.rm = TRUE),
    Q50 = quantile(int_rate, 0.50, na.rm = TRUE),
    Q75 = quantile(int_rate, 0.75, na.rm = TRUE),
    Min = min(int_rate, na.rm = TRUE),
    Max = max(int_rate, na.rm = TRUE)
  )

cat("Descriptive Stats of Interest Rates Separated by Loan Status)\n\n")
print(rate_stats)

```

```{r, echo=TRUE}
library(ggplot2)

df_count <-df_final %>%
  filter(int_rate >0) %>%
  mutate(int_rate_bin = floor(int_rate/2)*2) %>%
  filter(int_rate_bin >0) %>%
  count(int_rate_bin, loan_status)

ggplot(df_count, aes(x = int_rate_bin, y = n, color = loan_status)) + 
  geom_line(size = 1)+ 
  labs(x="Interest Rate (%)", y="Count", title = "Distribution of Interest Rates by Loan Status") +
  scale_color_manual(values = c("navy", "lightblue")) 

```

Based on the statistics, higher interest rates significantly increase the likelihood of a loan being charged off. Charged off loans have a mean interest rate of 15.8%, compared to 12.6% for fully paid loans, with a similar difference in the median (15.1% vs. 12.1%). Even at the lower end of the interest rate range, loans that default tend to have higher rates. As charged off loans have a higher standard deviation and higher 75th percentile, this suggests greater volatility and risk. This indicates that interest rates are a strong predictor of loan defaults.  

Similar to the statistics, the histogram shows interest rates having a strong influence on loan outcomes. Loans with a higher interest rate of around 15% and above are more likely to be charged off (in proportion to fully paid), indicating a greater risk of default at these levels. In contrast, lower interest rates between 10% to 12% are associated with a higher frequency of fully paid loans, suggesting that borrowers with lower rates are more likely to repay their loans in full. There is also a noticeable overlap in the 10% to 15% interest rate range, where fully paid and charged off loans occur. But as interest rates increase, the likelihood of a loan being charged off rises significantly. This suggests that interest rates are a key predictor of loan performance, with higher rates correlating strongly with loan defaults and lower rates with successful repayment.  

Overall, loans with higher interest rates are likelier to be charged off, making interest rates a strong influence of loan charge offs. 



#Question 2 and answers go here:

#Question 3 and answers go here:

#Question 4 and answers go here:


#Conclusion is here. 



