---
title: "Project one: Lending Club Data Analysis"
author: "Uyemaa Gantulga"
date: "10/15/2024"
output:
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
  pdf_document:
    toc: yes
    toc_depth: 3
---

### Setup chunk: Load libraries and set global options

```{r}
knitr::opts_chunk$set(echo = TRUE, include = TRUE)
library(knitr)
library(dplyr)
library(ggplot2)
library(kableExtra)

```

### Load the accepted and rejected loans dataset

```{r}
accepted_loans <- read.csv("C:\\Users\\uemaa\\Documents\\Data Science\\Project 1\\cleaned_accepted_2013_to_2018Q4.csv")
rejected_loans <- read.csv("C:\\Users\\uemaa\\Documents\\Data Science\\Project 1\\cleaned_rejected_2013_to_2018Q4.csv")
```

### Display column names of accepted loans dataset

```{r}
colnames(accepted_loans)
colnames(rejected_loans)
```

### Structure of accepted loans dataset

```{r}
str(accepted_loans)
```

### Structure of rejected loans dataset

```{r}
str(rejected_loans)
```

### Summary of accepted and rejected loans dataset

```{r}
summary(accepted_loans)
summary(rejected_loans)
```


## Question 4: Is there a relationship between employment length and the amount requested for rejected loans and how does it compare to the customers that had accepted loan




